---
layout: post
title: "Quantitative Ecotoxicology, Page 159, Example 4.6, Control Mortality"
date: 2013-03-11 23:23
author: Eduard Szöcs
published: true
status: published
draft: false
tags: QETXR R
---
```{r setup, echo=FALSE, cache=FALSE, message=FALSE}
options(scipen = 1, digits = 5)
require(knitcitations)
```
```{r echo=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(fig.height=6, fig.width=6)
```

This is example 4.6 on page 159 of [Quantitative Ecotoxicology](http://www.crcpress.com/product/isbn/9781439835647). It is about how to deal with control mortalities.


First we need the data:
```{r message=FALSE}
require(RCurl)
url <- getURL("https://raw.github.com/EDiLD/r-ed/master/quantitative_ecotoxicology/data/p160.csv",
ssl.verifypeer = FALSE, .opts=curlOptions(followlocation=TRUE))
NAP <- read.table(text = url, header = TRUE, sep = ";")
```
```{r}
head(NAP)
```

The data consists of number of dead animals (DEAD) from all animals (TOTAL) exposed to different concentrations (CONC).
First we create a new column with the proportion of dead animals:

```{r}
NAP$PROP <- NAP$DEAD / NAP$TOTAL
```

Here is a plot of the data. Note the use of `expression()` (greek letters in the axis labels).
```{r plot_raw}
plot(NAP$CONC, NAP$PROP, 
     pch = 16, 
     xlab = expression(paste('Concentration (', mu, 'g/L)')),
     ylab = 'Proportion Dead',
     main = 'Raw data')
```


### Control mortality

We can estimate the mean control mortality and the confidence interval for the mean using the `t.test` function:
```{r}
contr_m <- t.test(NAP$PROP[NAP$CONC == 0])
contr_m
```

These can be also easily extracted from the t.test object:

```{r}
## extract the values from t.test-object
# mean
contr_m$estimate
# CI
contr_m$conf.int
```

This gives nearly the same values as in the book. I don't know what the SAS option `OPTC` is doing or computing, however it seems like it is the mean +- CI for the control group.


### Abbott’s formula

We could adjust for control mortality using Abbott's formula:

$$p_c = \frac{p-p_0}{1-p_0}$$

with $p_c$ = the corrected, p = original and $p_0$ = control mortality.

The mean control mortality can be calculated as:
```{r}
d_control <- mean(NAP$PROP[NAP$CONC == 0])
d_control
```

And the corrected mortalities using Abbotts formula as:
```{r}
NAP$PROP_c <- (NAP$PROP - d_control) / (1 - d_control)
NAP$PROP_c
```


### Dose-Response-Models

#### Ignoring control mortality

As in the previous example we can fit a dose-response-model to this data using the `drc` package:
```{r message=FALSE, warning=FALSE}
require(drc)
mod1 <- drm(DEAD/TOTAL ~ CONC, weight = TOTAL, data = NAP, fct = LL.2(), type = 'binomial')
```

```{r plot_mod1}
plot(mod1, broken = TRUE, type = 'all', bp = 500, xt = seq(500,3000,500))
mtext('Dose-Response-Model - LL2.2', 3)
```


#### Using the corrected mortalities

We can also fit a model to the corrected mortalities `PROP_c`.

Abbotts correction resulted to some negative mortalities, therefore I set the control and all negative mortalities to zero:

```{r}
NAP$PROP_c[NAP$PROP_c < 0 | NAP$CONC == 0] <- 0
```

Then we fit a dose-response model:
```{r warning=FALSE}
mod2 <- drm(PROP_c ~ CONC, weights = TOTAL, data = NAP, fct = LL.2(), type = 'binomial')
```
 
```{r plot_mod2}
plot(mod2, broken = TRUE, type = 'all', bp = 500, xt = seq(500,3000,500))
mtext('Corrected mortalities - LL.2', 3)
```


#### A model without fixed lower limit

The two-parameter log-logistic model from above (`mod1`) performs quite good. However its lower limit is fixed to 0 and the upper limit to 1.
Since we have a small amount of control mortality we could check if a model with varying lower limit (will be estimated) makes sense.

Let's fit a three parameter log-logistic function, where the lower limit is an additional parameter:

```{r plot_mod3, warning=FALSE}
mod3 <- drm(PROP ~ CONC, weights = TOTAL, data = NAP, fct = LL.3u(), type = 'binomial')
plot(mod3, broken = TRUE, type = 'all', bp = 500, xt = seq(500,3000,500))
mtext('Free (estimated) lower limit - LL3.u', 3)
```

However looking at the summary we see that the lower limit (`c`) is basically zero.
```{r}
summary(mod3)
```

Since the lower limit (=control mortality) is so low we could also stick with the simpler model:

```{r}
anova(mod1, mod3)
```

suggest that there is no statistically significant better fit of the more complicated model.




All three considered models give nearly the same $LC_{50}$ around 2100:
```{r}
ED(mod1, 50, interval = 'delta')
ED(mod2, 50, interval = 'delta')
ED(mod3, 50, interval = 'delta')
```

