---
layout: post
title: "Web scraping chemical data with R"
date: 2014-07-09 19:38
author: Eduard Szöcs
published: true
status: process
draft: false
tags: R web-scraping
---

```{r setup, echo=FALSE, cache=FALSE, message=FALSE}
options(scipen = 1, digits = 5)
```
```{r echo=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(fig.height=6, fig.width=6, chache=TRUE)
require(methods)
```

I recently came across the problem to convert [CAS numbers](http://en.wikipedia.org/wiki/CAS_registry_number) into [SMILES](Simplified molecular-input line-entry system) and retrieve additional information about the compound.

The are several sources around the web that provide chemical informations, eg.
[PubChem](https://pubchem.ncbi.nlm.nih.gov/), [ChemSpider](www.chemspider.com) and the [Chemical Identifier Resolver](http://cactus.nci.nih.gov/chemical/structure).

I wrote up some functions to interact from R with these servers. You can find them in my [esmisc](https://github.com/EDiLD/esmisc) package:

```{r load, eval=FALSE}
install.packages('devtools')
require(devtools)
install_github('esmisc', 'EDiLD')
require(esmisc)
```

```{r load2, echo=FALSE, message=FALSE}
require(esmisc)
```

These functions are very crude and need some further development (if you want to improve, fork the package!), however, here's a short summary:

### Covert CAS to SMILES
Suppose we have some CAS numbers and want to convert them to SMILES:
```{r, cas}
casnr <- c("107-06-2", "107-13-1", "319-86-8")
```

##### Via Cactus
```{r, cactus_smiles}
cactus(casnr, output = 'smiles')
```

##### Via ChemSpider
Note, that ChemSpider requires a security token. To obtain a token please [register](http://www.chemspider.com/Register.aspx) at ChemSpider. 
```{r token, echo=FALSE}
token <- '37bf5e57-9091-42f5-9274-650a64398aaf'
```

```{r cs_smiles}
csid <- get_csid(casnr, token = token)
csid_to_smiles(csid, token)
```

##### Via PubChem
```{r pc_smiles}
cid <- get_cid(casnr)
cid_to_smiles(cid)
```



### Retrieve other data from CAS

All these web resources provide additional data. Here is an example retrieving the molecular weights:

##### Via cactus
```{r cactus_mw}
cactus(casnr, output = 'mw')
```

##### Via ChemSpider
```{r cs_mw}
csid_to_ext(csid, token)$MolecularWeight
```

##### Via Pubchem
```{r pc_mw}
cid_to_ext(cid)$mw
```
ChemSpider and PubChem return the same values, however the results from cactus are slightly different.


### Retrieve partitioning coefficients
Partition coefficients are another useful property. [LOGKOW](http://logkow.cisti.nrc.ca/logkow/intro.html) is a databank that contains experimental data, retrieved from the literature, on over 20,000 organic compounds. 

`get_kow()` extracts the 'Recommended values' for a given CAS:

```{r kow}
get_kow(casnr)
```

This function is very crude. For example, it returns only the first hit if multiple hits are found in the database - a better way would be to ask for user input, as we did the [taxize](https://github.com/ropensci/taxize) package.


### Outlook

Currently I have no time to extensively develop these functions. 
I would be happy if someone picks up this work - it's [fairly easy](https://help.github.com/articles/fork-a-repo#contributing-to-a-project): just fork the repo and start.

In future this could be turned into a [ROpenSci](http://ropensci.org/) package as it is within their scope.
