---
layout: post
title: "Quantitative Ecotoxicology, Page 223, Example 5.1, Using GLM"
date: 2015-01-19 22:08
author: Eduard Szöcs
published: false
status: process
draft: true
tags: QETXR R
---

```{r setup, echo=FALSE, cache=FALSE, message=FALSE}
options(scipen = 1, digits = 5)
require(knitcitations)
```
```{r echo=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(fig.height=6, fig.width=9)
```

[Previously](http://edild.github.io/quant-ecotox20/), I showed how to analyse the fish survival data using the arcsine square root transformation.
Warton & Hui (2011) demonstrated that *the arcsine transform should not be used in either circumstance*, but instead use Generalized Linear Models.
This post is about how to analyse example 5.1 on page 223 of [Quantitative Ecotoxicology](http://www.crcpress.com/product/isbn/9781439835647) using Generalised Linear Models.


### Introduction

Data of this type (*x out of n*) can be directly modelled using a [binomial distribution](http://en.wikipedia.org/wiki/Binomial_distribution).

<!--more-->

The binomial distribution is described by two parameters:

$$Bin(N, \pi)$$

where N is the number of fish and $\pi$ is the probability of survival.


Here I plotted the binomial distribution for ten fish at different probabilities ($\pi = 0.5, 0.75, 0.95, 0.05$), to give you a feeling how this distribution looks like at difference $\pi$:

```{r, plot_binomial, message=FALSE}
require(ggplot2)
require(reshape2)
df <- data.frame(k = 0:10, 
                 p1 = dbinom(0:10, size = 10, prob = 0.5),
                 p2 = dbinom(0:10, size = 10, prob = 0.75),
                 p3 = dbinom(0:10, size = 10, prob = 0.95),
                 p4 = dbinom(0:10, size = 10, prob = 0.05))
dfm <- melt(df, id = 'k')

ggplot(dfm, aes(x = factor(k), y = value, fill = variable)) +
  geom_bar(stat = 'identity', position = "dodge") +
  labs(x = expression(pi), y = expression(p(pi))) +
  scale_fill_discrete("Distribution", 
                    labels = c("Bin(10, 0.5)", "Bin(10, 0.75)", "Bin(10, 0.95)", "Bin(10, 0.05)")) +
  theme_bw()
```



### The Model

We can now build a model, where the probability of success is a function of treatment:

$$  y \sim Bin(N, \pi) \\
  logit~(\pi) = \alpha + \beta x \\
  var(y) =  \pi (1 - \pi) / N $$
  
  
Let's explain this a little bit:

$$ y \sim Bin(N, \pi) $$, basically says: We assume that the number of dead fish ($y$) binomially distributed, where N = exposed animals and $\pi$ i is the probability of survival, which together give the expected number of surviving fish ($E(y) = N \times \pi$).

$$ logit~(\pi) = \alpha + \beta x $$, basically says: We are modelling the probability of survival as function of treatment (x) [note the right-hand side of the formula is similar to linear regression]. However, we need to ensure that $$0 < \pi < 1$$, and therefore the relationship is on the logit scale. The estimated parameters ($\beta) are directly interpretable as changes in log odds between treatments.

$$ var(y) =  \pi (1 - \pi) / N $$, says that variance of the binomial distribution
is a quadratic function of the mean. Note, that in linear regression we assume a constant variance.



### How to do it in R?

First I read again the data into R and prepare it:

```{r}
df <- read.table(header = TRUE, text = 'conc A B C D
0 1 1 0.9 0.9
32 0.8 0.8 1 0.8
64 0.9 1 1 1 
128 0.9 0.9 0.8 1
256 0.7 0.9 1 0.5
512 0.4 0.3 0.4 0.2')
require(reshape2)
dfm <- melt(df, id.vars = 'conc', value.name = 'y', variable.name = 'tank')
dfm$conc <- factor(dfm$conc)
```


The compute a binomial GLM in R we can use the `glm()` function:

```{r}
modglm <- glm(y ~ conc , family = binomial(link = 'logit'), data = dfm, 
              weights = rep(10, nrow(dfm)))
```

Note, that I use the `weights` argument to specify that in each tank we had 10 fish (N in the above formulas).
The summary gives the estimated parameters:

```{r}
summary(modglm)
```

The estimates for `(Intercept)` are the log odds to survive in the treatment:


```{r}
# mean proportion surviving in control
p_0 <- mean(dfm$y[dfm$conc == 0])
# log odds
(logodd_0 <- log(p_0 / (1 - p_0)))
```

Similar the log odds for the highest treatment:
```{r}
# mean proportion surviving in control
p_512 <- mean(dfm$y[dfm$conc == 512])
# log odds
(logodd_512 <- log(p_512 / (1 - p_512)))
```

The estimate for `conc512` (-3.675) gives you the difference in the log odds, as can be seen here:
```{r}
logodd_512 - logodd_0
```

Note, that this kind of interpretation is not possible with the arcsine transformation.

#### Hypothesis tests
Similar to the previous post we can perform an F-Test:

```{r}
drop1(modglm, test = 'F')
```

Or do multiple comparisons:

```{r, message=FALSE}
require(multcomp)
summary(glht(modglm, linfct = mcp(conc = 'Dunnett')))
```




### Which one is better?
I would say GLMs! 
They have greater power (Warton & Hui, 2011), are simpler to interpret and are readily available in most software packages.


#### References

* Warton, D. I., & Hui, F. K. (2011). The arcsine is asinine: the analysis of proportions in ecology. Ecology, 92(1), 3-10.
