---
layout: post
title: webchem v0.0.3 is out!
author: "Eduard Szöcs"
date: "September 25, 2015"
published: true
status: published
draft: false
tags: R ecotoxicology web-scraping ropensci
---

A new version of the webchem package is available on [CRAN](https://cran.r-project.org/web/packages/webchem/index.html).
Don't forget to update!

This version add some functionality especially useful for ecotoxicologists and fixes some minor bugs.

Thanks to [Johannes Ranke](https://github.com/jranke) for fixing a typo with Alan Wood's database:
The old function name was `allanwood()` - this function is now deprecated and the new one is `alanwood()`.

If you want to contribute - please help me to improve webchem!
Either by creating a [pull request](https://github.com/ropensci/webchem/blob/master/CONTRIBUTING.md) or by [raising an issue](https://github.com/ropensci/webchem/issues) if you find a bug or some functionality is missing or by any other contribution.


### New data sources

#### SRC PHYSPROP

You can now query [SRC physprop database](http://www.srcinc.com/what-we-do/environmental/scientific-databases.html), which contains names and physical properties for over 41,000 chemicals. 

Currently, you can only search for CAS numbers (here for Diuron):

```{r, message = FALSE}
library(webchem)
physprop('330-54-1')
```


#### ETOX

[ETOX: Information System Ecotoxicology and Environmental Quality Targets](http://webetox.uba.de/webETOX/index.do?language=en) is a nice database which holds information on substances (a lot of synonyms in there), tests and quality targets.

To use it we follow the common workflow: *Search ID -> Find data with this ID*.

* First we need the ETOX-ID:
```{r, message = FALSE}
id <- get_etoxid('Triclosan')
id
```

With this ID we can query **substance information**:

```{r, message = FALSE}
etox_basic(id)
```

Or **quality targets**:

```{r, message = FALSE}
targets <- etox_targets(id)
targets[ , c('Substance', 'Country_or_Region', 'Designation', 'Value_Target_LR', 'Unit')]
```

And (similar to the EPA AQUIRE database) results of **toxicity tests**:

```{r, message = FALSE}
tests <- etox_tests(id)
tests[ , c('Organism', 'Effect', 'Duration', 'Time_Unit','Endpoint', 'Value', 'Unit')]
```


#### PPDB
The [PPDB : Pesticides Properties DataBase](http://sitem.herts.ac.uk/aeru/iupac/index.htm) is another useful data source for ecotoxicologists. 

Currently, you can query the PPDB only using a CAS number and the function will definitively some improvements in future.

```{r, message = FALSE}
out <- ppdb_query('330-54-1')
```

The output is enormous, so I print here only a small part, e.g. the EU member states where
CAS 330-54-1 (=Diuron) is approved for use.

```{r}
out[[2]]
```


