---
title: "New version of esmisc"
author: "Eduard Szöcs"
date: "2017-01-10 20:00"
layout: post
published: false
status: published
tags: R 
draft: true
---
<img src="http://vg03.met.vgwort.de/na/297d36f243e3462ab6fbb1d07fa0c442" width="1" height="1" alt="">

```{r setup, echo=FALSE, cache=FALSE, message=FALSE}
options(scipen = 1, digits = 5)
```

```{r knitr_setup, echo=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(fig.height = 6, fig.width = 8, message = FALSE, 
               echo = TRUE, cache = TRUE)
```

A user of `read_regnie()` [(a function to read precipitation data from DWD)](http://edild.github.io/regnie/) pointed me to some problems with some raster files that throw an error.
It turned out that the function could not read rasters with a precipitation greater than 99 mm.
The bug was fixed and also the function is now 4.5x faster than the old code.

Check out the new version of `esmisc` on [github](https://github.com/EDiLD/esmisc).

 `esmisc` is now also available from [CRAN](https://CRAN.R-project.org/package=esmisc).
