---
title: "Regulatory Acceptable Concentrations (RAC)"
author: "Eduard Szöcs"
date: "2016-12-05 12:00"
layout: post
published: false
status: published
tags: R ecotoxicology
draft: false
---
<img src="http://vg03.met.vgwort.de/na/b4a181848e034cc0a5a6b61bd416b041" width="1" height="1" alt="">

```{r setup, echo=FALSE, cache=FALSE, message=FALSE}
options(scipen = 1, digits = 5)
```

```{r knitr_setup, echo=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(fig.height = 6, fig.width = 8, message = FALSE, 
               echo = TRUE, cache = TRUE)
```

## Intro

The German Environment Agency (UBA) recently published a list of regulatory acceptable concentrations (RAC) for 108 pesticides [[Link](https://webetox.uba.de/webETOX/public/basics/literatur.do?id=24559)].
I used it already in a previous project (paper is currently under review) and here I show how we can use `R` to:

1. Download the list
2. Digitize the pdf
3. Retrieve additional data
3. Create summary statistics to explore the RACs

I will use two [ROpenSci](https://ropensci.org/) packages for this tasks:

1. [tabulizer](https://github.com/ropenscilabs/tabulizer) to digitize the pdf and
2. my [webchem package](https://github.com/ropensci/webchem) to retrieve information on the pesticides.


## Download the data
First, we need to get the RAC-list from ETOX.

Here I download the file to a temporary file:

```{r download}
rac_original <- tempfile()
download.file('https://webetox.uba.de/webETOX/public/basics/literatur/download.do?id=284', 
                destfile = rac_original)
```


## Digitize the data

Next we need to get the pdf into a tabular format.
I use [tabulizer](https://github.com/ropenscilabs/tabulizer) for this task.
We digitize the pdf table using `extract_tables()`:

```{r extract_tables}
library("tabulizer")
rac_list <- extract_tables(rac_original)
str(rac_list)
```

This gives two lists with matrices because there are two pages.
We combine both 

```{r clean_tables1}
rac_list <- do.call(rbind, rac_list)
str(rac_list)
```
We see that there are 109 instead of 108 lines.
Inspecting the raw pdf and the table we see that there is an addtional for Diuron with 2 CAS numbers and no indentifer in the first column.
I simply delete this row:

```{r clean_tables2}
rac_list <- rac_list[!rac_list[ ,1] == '', ]
```


## Clean the data
Some more cleaning is needed before we can use this data efficiently:

1. Delete unused columns
2. Coerce to `data.frame`

```{r clean_tables3}
# keep only selected columns and make data.frame
rac <- data.frame(rac_list[ , c(1, 2, 4, 6)], stringsAsFactors = FALSE)
head(rac)
```

3. split columns missed by tabulizer.
```{r clean_tables4, warning = FALSE}
# split columns X2 and X3
library("tidyr")
rac <- separate(data = rac, col = X2, into = c("name", "type"), sep = -3)
rac <- separate(data = rac, col = X3, into = c("CAS", "date"), sep = " ")
# remove date 
rac$date <- NULL
head(rac)
```


4. Check CAS numbers
```{r clean_tables5}
# this is needed to fix an encoding problem
rac$CAS <- gsub('‐', '-', rac$CAS)
library(webchem)
# check if all are valid cas numbers
# Note in a future release of webchem is.cas will be vectorized!
which(vapply(rac$CAS, is.cas, FUN.VALUE = TRUE) == FALSE)
```
One CAS is wrong (date in stead of CAS), which I set to NA
```{r}
rac$CAS[39] <-  NA
```


5. coerce rac to `numeric` (currently it is text)
```{r clean_tables6}
rac$rac <- as.numeric(gsub(",", ".", rac$X4))
```

6. final cleanup

```{r clean_tables7}
# select columns
rac <- rac[ , c('name', 'CAS', 'rac')]
# remove substances without rac
rac <- rac[!is.na(rac$rac), ]
# trim whitespaces (leading & trailling)
rac$name <- gsub("^\\s+|\\s+$", "", rac$name)
# set cas for dodin to 2439-10-3
rac$CAS[rac$name == 'Dodin'] <- '2439-10-3'
head(rac)
# # save
# write.table(rac, file = '../files/rac.csv', sep = ';', row.names = FALSE)
```

You can download the cleaned file [here](http://edild.github.io/files/rac.csv).


## Retrieve additional information

I search for additional information, like the activity, in [Alan Wood's Compendium of Pesticide Common Names](http://www.alanwood.net/pesticides/):

```{r query_aw}
aw_results <- aw_query(rac$CAS, type = 'cas')
```
This worked very well and we found information for all substances except for a Prothioconazole metabolite

```{r check_aw}
# Check for which CAS no data (=NA) was found
which(vapply(aw_results, function(y) length(y) == 1, TRUE))
```

I remove this from the results and the rac list.

```{r fix_aw}
rac <- rac[!rac$CAS == '120983-64-4', ]
aw_results <- aw_results[!names(aw_results) == '120983-64-4']
```

Finally, I extract the subactivity and add it to the rac table:

```{r exrtact_aw}
rac$subactivity <- vapply(aw_results, function(y) y[['subactivity']][1], 'x')
```

From this I extract the activity and build 4 groups (insecticides, herbicides, fungicides and others):

```{r fix_aw2}
rac$type <- gsub('.* (.*)$', '\\1', rac$subactivity)
# fix
rac$type[rac$type == 'insecticidesmolluscicides'] <- 'insecticides'
# build groups
rac$type <- ifelse(!rac$type %in% c('herbicides', 'fungicides', 'insecticides'),
                   'other',
                   rac$type)

```


Next, I search the [Physprop database](http://esc.syrres.com/fatepointer/webprop.asp?CAS=135410207) for $K_{OW}$ values.

```{r pp_query, warning=FALSE}
pp_results <- pp_query(rac$CAS)
# Check for which CAS no data was found
which(vapply(pp_results, function(y) length(y) == 1, TRUE))
```
We could not find data for 20 substances and I set their value to `NA`.

 
```{r pp_extract}
rac$kow <- sapply(pp_results, function(y) {
    # substances without data: NA
    if (length(y) == 1 && is.na(y))
      return(NA)
    out <- y$prop$value[y$prop$variable == 'Log P (octanol-water)']
    # substances without KOW: NA
    ifelse(length(out) == 0, NA, out)
  })
```



## Summarize data

All the code above was just done to create few summary statistics...

Here the RACs splitted by type:
```{r plot_type}
library('ggplot2')
ggplot(rac, aes(y = rac, x = type)) +
  geom_boxplot(fill = 'grey75') +
  geom_jitter(width = 0.2) +
  scale_y_log10() +
  theme_bw() +
  labs(x = '', y = 'RAC [ug/L]')
```

We clearly see that insecticides have much lower RACs than the other groups of pesticides.
However, there is no strong relationship between the $log~K_{OW}$ and the RAC for the different types.

```{r plot_kow, warning=FALSE}
ggplot(rac, aes(y = rac, x = kow, col = type)) +
  geom_point() +
  scale_y_log10() +
  theme_bw() +
  facet_wrap(~type) +
  labs(x = expression('log'~K[OW]), y = 'RAC [ug/L]')
```






