---
title: "Experimental design for dose-reponse-experiments - a simulation"
author: "Eduard Szöcs"
date: "2016-6-14 20:00"
output: pdf_document
layout: post
published: true
status: published
tags: R ecotoxicology
draft: false
---

```{r setup, echo=FALSE, cache=FALSE, message=FALSE}
options(scipen = 1, digits = 5)
```
```{r echo=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(fig.height = 6, fig.width = 6, chache = TRUE, message = FALSE, echo = TRUE, cache = TRUE)
require(methods)
```


## Introduction

Ecotoxicologist face often the problem how to allocate resource when conducting a dose-response experiment to determine the $EC_{50}$ of species towards a substance.

*How many concentrations should I test?* *How many replicates should I run per concentration?*
*How many animals should I use per replicate?*

Often the answers are constrained by laboratory resources (e.g. How many beakers are available?) or the amount of work per replicate (e.g. How much time do I need to prepare I replicate?).

One way to address such questions is to run simulations of different experimental setups and to compare their performance.
In this post I will demonstrate how to run such simulations.


## Simulation setup

First of all we need a algorithm that generates data where we know the properties of.
I will simulate data from a logistic regression with a binomial response variable

$y \sim Bin(N, p)$

Here `N` is the total number of animals and `p` is the probability of an animal to survive.
I simulate the probability `p` as function of the concentration `x`:

$logit(p) = \alpha + \beta~x$

where $\alpha$ and $\beta$ are the parameters of the relationship and the $EC_{50}$ is given by $- \frac{\alpha}{\beta}$.



## Simulation scenarious

`x` values were evenly spaced in $0 \le x \le 1$, with varying numbers of treatments: 

$n_{x} = \{4,6,8,10,12\}$.

I will also modify the number of replicates per treatment:

$n_{rep} = \{1, 3, 5, 7, 9\}$

And the number of animals per replicate:

$N = \{3, 6, 9, 12, 15, 18, 21, 24, 27, 30\}$.

I chose $\alpha = 5$ and $\beta = -10$, giving a LC50 of 0.5 and a decent slope.

First we need a function that generates data, according to the above described setup:

```{r datagen}
# Data generation function
#' -param alpha intercept
#' -param beta slope  
#' -param n_x number of treatments
#' -param n_rep number of replicates per treatment
#' -param N number of animals per replicate
#' -param n_sim number of simulation runs
#' -return a list with three components:
#'  x: x-values, y = a matrix, each column is one simulation run, N = number of animals
data_gen <- function(alpha = 5, beta = -10, n_x = 6, 
                    n_rep = 3,  N = 10, n_sim = 1){
  # evenly space x-values
  x <- seq(0, 1, length.out = n_x) 
  # inv-logit
  p <-  1 / (1 + exp(-(alpha + beta*x)))  
  # sample from binomial
  y <-  replicate(n_sim, rbinom(n_x * n_rep, N, p)) 
  return(list(x = rep(x, n_rep), y = y, N = N))
}
```

The arguments of this functions are all described above, let's see how the generated data looks like:

Here I generate data with the default set-up ($\alpha = 5$, $\beta = -10$, $n_x = 6$, $n_{rep} = 3$, $N = 10$ and $n_{sim}=1$)

```{r datagen_ex1}
sim1 <- data_gen()
plot(sim1$x, sim1$y[, 1], xlab = 'Conc.', ylab = 'Animals survived')
```

We can fit a log-logistic dose-response model (DRM) to this data using the `drc` package to determine the LC50

```{r drc_ex1}
library(drc)
mod <- drm(y/sim1$N ~ x, 
    weights = rep(sim1$N, length(x)),
    data = data.frame(x = sim1$x, y = sim1$y[, 1]), 
    fct = LL2.2(), 
    type = "binomial")
plot(mod)
ED(mod, 50, interval = 'fls')
```

We see that the simulated data corresponds to the $EC_{50}$ we aimed to simulate (=0.5).

As we will have a lot of simulation scenarios ($5 \cdot 5 \cdot 10~=~250$) we need a function to analyse each generated dataset. 
I am only interested in the deviation from the specified $EC_{50}$ value and therefore use the absolute-error (AE) from 0.5 as benchmark.
Here is a function that fits a log-logistic drm to given simulated data and returns the AE from the specified value of 0.5:

```{r data_ana}
#' analysis function
#' -param z a list as returned by data_gen()
data_ana <- function(z) {
  ana <- function(x, y, N) {
    # model
    mod <- try(drm(y/N ~ x, 
                   weights = rep(N, length(x)),
                   data = data.frame(x = x, y = y), 
                   fct = LL2.2(), type = "binomial"))
    if (inherits(mod, "try-error")) {
      bias <- NA
    } else {
      # LC50
      LC50 <- try(exp(ED(mod, 50, interval = 'none', 
                         display = FALSE)[1, 1]), silent = TRUE)
      if (inherits(LC50, "try-error")) {
        bias <- NA
      } else {
        # AE
        bias <- abs(LC50 - 0.5)
      }
    }
    return(bias)
  }
  out <- apply(z$y, 2, ana, x = z$x, N = z$N)
  return(out)
}
```

This function fits a DRM to a given dataset and returns the absolute deviation from 0.5.
I wrapped some calls into `try()` because for some data we might fails to fit a model (because of convergence problem , complete separation etc...).

Let's see the function in action with the `sim1` object from above:

```{r ae}
data_ana(sim1)
```

and a cross-check:
```{r ae2}
abs(ED(mod, 50, interval = 'fls', display = FALSE)[1] - 0.5)
```



OK, now we have the parts we need - let's get to work.
First I define the scenarios I want to run through:

```{r scenarious}
scenarious <- expand.grid(N = seq(3, 30, 3),
                          n_rep = seq(1,9 , 2),
                          n_x = seq(4, 12, 2))
head(scenarious)
```

Each row corresponds to a specific scenario.
I will now run through all scenarios, generate 100 datasets from each and calculated the AE of each:

[Important note: This code block takes some time to run! I actually run it on a Amazon EC2 instance...]
```{r sim, warning=FALSE, eval=FALSE}
ae <- NULL
# for each scenario
for (i in seq_len(nrow(scenarious))) {
  message('scenario: ', i)
  take <- scenarious[i, ]
  sims <- data_gen(n_rep = take$n_rep, N = take$N, 
                n_x = take$n_x, n_sim = 100)
  ae[[i]] <- data_ana(sims)
}
```

```{r cache_file, include=FALSE, eval=FALSE, echo=FALSE}
saveRDS(ae, "/home/edisz/Documents/Uni/Projects/edild.github.com//files/ae.rds")
```

```{r read_cache, echo=FALSE}
ae <- readRDS("/home/edisz/Documents/Uni/Projects/edild.github.com//files/ae.rds")
```


The bias of each of the 100 generated datasets for each scenario are stored in `ae`.
Next I compute the mean of the AE for each scenario and add it the the `scenarious` data.frame:

```{r mean_ae}
scenarious$mean_ae <- sapply(ae, mean, na.rm = TRUE)
```

## Results
Let's look at the results. Because we have 3 factors possibly affecting the error in $EC_{50}$ determination if will use a faceted plot:

```{r plot_results, fig.width=8, fig.height=8}
library(ggplot2)
ggplot(scenarious, aes(x = N, y = mean_ae)) +
  geom_line() +
  geom_point() +
  facet_grid(n_rep ~ n_x, labeller = label_both) +
  theme_bw() +
  labs(y = 'Mean absolute error', x = 'N')
```

We see that with increasing $N$, $n_{rep}$ and $n_x$ the absolute error decreases and levels off at something around 0.02.

Having only 4 tested concentrations (left-most column) is not feasible, as the AE never reaches this low level.
Also if you have only one replicate you need to have either lots of concentrations or animals (or both).
However, having lots of replicates, animals and concentrations does not add much and is possibly a waste of resources.

The displayed plot matrix can help to decide how to allocate your resources.
This simulation is of course only valid in the specified scenarios, but could be easily modified to other scenarios.





